---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Abishek Nippani"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

In this analysis, I look at **median gross rent across counties in New York**, and
ask which communities we measure most and least reliably. I have chosen to look at
the median gross rent people pay in New York as the state has the highest average
rental costs in the world. I'm curious to see how this average changes across different
counties and regions of the state.


```{r}
my_state <- "36"        # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

The dataframe has 62 rows, which matches the number of counties in the state New York.


```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


Hamilton County has the largest margin of error relative to its estimate. The raw
MOE is 95, while the MOE percentage relative to its estimate is 11.04.


## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


51 counties fall into high confidence, 10 into moderate confidence and 1 into low confidence.

## Is the margin of error bigger in small places?

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```

Based on the size of the margin of error relative to its estimate (moe_pct), three buckets of reliability
were created: *high confidence* (moe_pct<5), *moderate confidence* (moe_pct<10) and 
*low confidence* (moe_pct>=10). 51 counties fall into high confidence, 10 into moderate
and 1 into low confidence. 

A clear pattern can be observed across the three buckets: as me move from high to moderate
to low confidence, the average population of these counties also falls. 

Specifically, the average population of high confidence counties is 381,479,
41,178 for moderate confidence counties, and 5,102 for the low confidence one.

The data seems to suggest that smaller counties may be more likely to have higher
margin of error relative to the underlying estimate.


## What this means for policy

When looking at county-wise median gross rental payments in the state of New York, 
we know the least about Hamilton County. This is based on the size of the margin 
of error relative to its corresponding estimate. Hamilton County has both the 
highest margin of error in absolute terms and the highest margin of error relative 
to the estimate: an MOE of 95 and an MOE of 11% relative to the estimate. 

This is a function of its small population size and the sampling strategy used 
in the ACS. Data for the present analysis is drawn from ACS, which is based on 
sample surveys. These surveys draw smaller samples from less populous counties 
and larger samples from more populous ones. Counties such as Putnam County or 
Greene County have an MOE of 80 and 79, respectively, but their MOE percentages 
are 4.4 and 7.96, respectively. While their MOEs are 84% (Putnam) and 83% (Greene) 
of Hamilton's, their MOE percentages are only 40% and 72% of Hamilton. 
This is because their respective populations are substantially bigger than 
Hamilton's 5,102 people -- with Putnam at 97,988 and Greene at 48,554. 

If the state of New York or federal agencies were to use data from the ACS to 
devise policies related to rental housing, then it could be very inaccurate for 
smaller counties such as Hamilton County. Specifically, if the estimate reflects the 
lower end of the true rent being paid in Hamilton, then policies would end up 
underserving people who pay higher rents. Thus, before any action is taken, it 
would be advisable to conduct specific surveys in smaller counties or triangulate 
the estimate with county-specific administrative data.

